neonstore: provenance-aware persistent storage for NEON data

### 2021-03-29 paper/paper.Rmd
:pen:

## Rationale and package design

@@

NEON data products are distributed using a highly atomized file design: with each data product from either both observation systems (OS) and instrument systems (IS) divided into separate files by site and month.  Products generated by the airborne observation platform (AOP) are typically divided by type, flight date, flight line, and grid cell.  This atomized approach facilitates sub-setting and data transfer, because users can identify ahead of time which products, locations, and time ranges they need. As some NEON data products can be quite large (NEON's complete data catalog to date is already in excess of 1 petabyte, primarily due to the size of many AOP products), being able to download or read only the files of interest is already important to save bandwidth or computer memory, or disk space. This approach also facilitates data provenance and reproducibility, because it means that an update or correction to data collected at a particular site and time results only in a change to a single file and not the whole data product. However, the highly atomized structure of NEON's data can also become a barrier to entry.  Most analyses will inevitably require multiple files to be combined or stacked, e.g. to create a unified record for a product across many months or years of sampling or across many sites. In addition to releasing new data files each month, NEON will occasionally update previously released data files.  More confusingly, NEON frequently updates the timestamp associated with a file even when the file has not changed. Typically these events are associated with the update of a separate but related file.

Any analysis is only as good as as the data supporting it.  Because NEON data is subject to both revisions (where previously released data are corrected) and new releases (where more recently collected data is made public), researchers will seek to update previous analyses with corrected data and new observations.  `neonstore` enables scripted workflows through which such updates can occur automatically by re-running the same script.  However, when new or revised data significantly alter results, it is also important to be able to pinpoint precisely what has changed and why.  These elements creates five distinct tasks for managing NEON data provenance: (1) The ability to determine which of each of the individual raw data files were used in an analysis (2) The ability to append new data as it is collected and released (3) the ability to transparently replace existing data when corrections are published, (4) the ability to avoid re-downloading or re-stacking data that has been previously processed, (5) the ability to freeze the data used in a given analysis, so that results can reliably be repeated without accidentally including revised or newer data products.

@@

@@

The location of the local store can be configured depending on the context.  The default behavior is to rely on a user-specific location appropriate for the operating system, e.g. `~/.local/share/neonstore` on Linux, `C:\Users\<username>\AppData\Local\neonstore` on Windows, or `~/Library/Application Support/neonstore` on Mac OS X, as determined by the `rappdirs` R package which provides appropriate modern defaults [@rappdirs].  This approach works well for a single user setup such as a laptop, where a user can share the store across multiple projects in different directories.  On a server system that is accessed by multiple users, an lower-level location that permits all users to read and write files to the store may be preferable.  Because `neonstore` only writes raw NEON data to the store, there is no risk of overwriting work of other users, while the shared storage can reduce consumption of redundant download and storage.  In yet other cases, a user may prefer a path that is specific to a single project.

@@

@@

To download this data, `neon_download()` must make many API calls.  It first uses `neon_sites()` to reference all sites containing this data product.  Then for each site, for each month for which that site collected data, the function will make an API call to the `/data` endpoint (using the `neonstore` function, `neon_data()`), which returns a table of files found for that product at the requested site and month range.  The  table from `neon_data()` indicates 4 values: the file name, the MD5 hash of the data file (confusingly labeled as the crc32 sum, though CRC32 hashes are used only for AOP data products.  Note that MD5 hashes are length-32 character strings, while the CRC32 hashes are 8 characters long), the file size (in bytes -- encoded as a character because some file sizes are too long to be represented as an integer), and a download URL.  The download URL embeds a temporary access token which expires after an hour, so users must complete all the downloads before this time expires.  The access string for each file is unique, thus it is (at the time of writing) impossible to download any given data file without making a corresponding API request, even if the precise file name and address is already known.  For the most part, this issue is invisible to `neonstore` users, as `neon_download()` calls automatically pass all download URLs to the download step.

```{r}
neon_download("DP1.10003.001", 
               start_date = "2018-01-01", 
               end_date = "2019-01-01",
               site = "YELL")

 ## Advanced use: filter for a particular table in the product
 neon_download(product = "DP1.10003.001",
               start_date = "2018-01-01",
               end_date = "2019-01-01",
               site = "YELL",
               file_regex = ".*brd_countdata.*\\.csv")
```

@@

Even adding a handful of products can generate thousands or hundreds of thousands of files in a local store. Surprisingly, these can even exceed the capacity of common bash commands like `ls *.csv`.  While a directory structure could mitigate this issue, `neonstore` sidesteps this issue through a convenient R-based interface.  The function `neon_index()` will list all recognized files found in the local store as a `data.frame`.  `neon_index()` uses a sophisticated filename parser, `neon_parse_filenames()`, that understands all documented file naming conventions listed at <https://data.neonscience.org/file-naming-conventions>, and quite a few that are not yet listed (internal NEON communication).  The most common of these are displayed in the table:

@@

`neon_index()` takes optional arguments such as `product`, `start_date`, `end_date`, and `site` to allow for convenient sub-setting.  Because most observational and instrument-based products are divided by site and month, this is still not a convenient way to browse the data.  The helper function, `neon_store()`, summarizes products that all share the same file description and can thus be easily stacked.  Like `neon_index()`, optional arguments can filter the results to specific products, dates, and sites.

```{r}
neon_store()
```

In this example, we recognize several tables corresponding to the NEON bird survey, product `DP01.10003.001` (all prefixed with `brd_`).  We can read in all the individual files

>>>>

# Rationale and package design

@@

`neonstore` focus on provenance emphasizes being able to trace results back precisely to individual raw data files provided by the NEON API.
NEON data products are distributed using a highly atomized file design: with each data product from either both observation systems (OS) and instrument systems (IS) divided into separate files by site and month.  Products generated by the airborne observation platform (AOP) are typically divided by type, flight date, flight line, and grid cell.  This atomized approach facilitates sub-setting and data transfer, because users can identify ahead of time which products, locations, and time ranges they need. As some NEON data products can be quite large (NEON's complete data catalog to date is already in excess of 1 petabyte, primarily due to the size of many AOP products), being able to download or read only the files of interest is already important to save bandwidth or computer memory, or disk space. This approach also facilitates data provenance and reproducibility, because it means that an update or correction to data collected at a particular site and time results only in a change to a single file and not the whole data product. However, the highly atomized structure of NEON's data can also become a barrier to entry.  Most analyses will inevitably require multiple files to be combined or stacked, e.g. to create a unified record for a product across many months or years of sampling or across many sites. `neonstore` provides functions to stack data products that have been atomized into separate files by month, site or sensor position into single `data.frame` objects, store stacked tables in a local, high-performance database, and track individual data rows back to their raw data files.

Managing provenance of NEON data means dealing with the addition of new data, corrections published to old data, and changes to file names that do not represent changes in underlying data.  Any analysis is only as good as as the data supporting it.  Because NEON data is subject to both revisions (where previously released data are corrected) and new releases (where more recently collected data is made public), researchers will seek to update previous analyses with corrected data and new observations.  `neonstore` enables scripted workflows through which such updates can occur automatically by re-running the same script. More confusingly, NEON frequently updates the timestamp associated with a file even when the file has not changed. Typically these events are associated with the update of a separate but related file. In such cases, `neonstore` can avoid downloading and importing these unchanged data, ensuring that automated workflows do not run unnecessarily.   When new or revised data significantly alter results, it is also important to be able to pinpoint precisely what has changed and why.  These elements creates five distinct tasks for managing NEON data provenance: (1) The ability to determine which of each of the individual raw data files were used in an analysis (2) The ability to append new data as it is collected and released (3) the ability to transparently replace existing data when corrections are published, (4) the ability to avoid re-downloading or re-stacking data that has been previously processed, (5) the ability to freeze the data used in a given analysis, so that results can reliably be repeated without accidentally including revised or newer data products.

@@

## Performance

`neonstore` is also concerned with performance of research workflows based on NEON data.  Native operations in R are based in memory, making it difficult to work with objects that are larger than RAM.  Researchers seeking to take full advantage of NEON data collected across all available sites and years can quickly run ub against the limits of data which can be held and processed in working memory.  The R language also supports interfaces with relational databases [@dbi], with the widely used package `dplyr` capable of seamlessly tanslating common functions (e.g. `filter`, `select`, `mutate`, `group_by`, `summarise`, `inner_join`, etc) into SQL queries that can be run by external database servers [@dplyr; @dbplyr].  This allows users to perform common tasks on data much larger than would fit into working RAM.  Traditional relational databases such as MariaDB and Postgres use a server-client model with row-oriented data structures, suitable for concurrent read and write operations by many simultaneous client processes acccessing a central database (e.g. many client banks simultaneously updating a central registry of accounts, see ACID).  Modern data science workflows frequently benefit more from column-oriented data structures where simultaneous operations (ACID transactions) are not required. This makes it possible to both eliminate the requirement of separately installing and running a database server while also opening the door to substantial performance improvements beyond the reach of traditional relational database clients.  `duckdb` is a leading example of such a columnar database [@duckdb], with seamless support for R integration and superior performance benchmarks to both traditional relational databases and in-memory operations in R using `dplyr`.

`neonstore` allows users to easily import tabular NEON data products (including `.h5` based eddy covariance products) into a local `duckdb` database using the function `neon_store()`.  If desired, users can instead use their own existing databases by providing the appropriate `DBI::dbConnection` object to the `db` argument of `neon_store()`. `neon_store()` stacks corresponding individual raw tabular data files by month, site, and (for IS data) sensor position into individual database tables.  `neon_store()` is capable of detecting revisions to previously released data products, removing outdated rows and replacing them with updated rows without accidental duplication of data.  Data revisions are indicated by downloading files which share the same data product, site, month and sensor position file name as files already present in the local store, but have an updated timestamp and differing content hash. `neon_store()` provides a warning whenever existing data tables are updated, as this may require previous analyses to be re-run.  Because sensor data does not indicate the site or position information in individual data files (this is provided only through the file name) `neonstore` stacking functions add additional columns for these variables automatically.  Additionally, a `file` column is always included indicating the raw file source for each row.  Many NEON data products come in two types: a "basic" format and an "expanded" format which may include additional columns providing data fields that were not included in the original product specification.  NEON does not always use globally unique names for table descriptions, particularly for metadata tables such as `sensor_positions`. To avoid ambiguities created by these issues, database table names are formed using the table description name, the product type, and the product number, e.g. `brd_countdata-basic-DP1.10003.001`, rather than the description component alone, `brd_countdata`.

RStudio users can bring up an interactive list of available tables in the RStudio "Connections" pane using the function `neon_pane()` [Fig #].

@@

The location of the local store can be configured depending on the context.  The default behavior is to rely on a user-specific location appropriate for the operating system, e.g. `~/.local/share/R/neonstore` on Linux, `C:\Users\<username>\AppData\Local\R\neonstore` on Windows, or `~/Library/Application Support/R/neonstore` on Mac OS X, as determined by `tools::R_user_dir()` which provides appropriate modern defaults [see @rappdirs].  This approach works well for a single user setup such as a laptop, where a user can share the store across multiple projects in different directories.  On a server system that is accessed by multiple users, a shared location that permits all users to read and write files to the store may be preferable.  Because `neonstore` only writes raw NEON data to the store, there is no risk of overwriting work of other users, while the shared storage can reduce consumption of redundant download and storage.  In yet other cases, a user may prefer a path that is specific to a single project.

@@

The local `neonstore` database is written in the same user and OS specific location by default, but is configured using a separate environmental variable, `NEONSTORE_DB`, or optional argument to appropriate functions,  `db_dir`.  Because simultaneous write-access is not supported by the default high-performance database engine, `duckdb`, in multi-user environments it may be advantageous to share a location for raw data files by setting `NEONSTORE_HOME` while allowing users to create individual local databases in their home directories by leaving `NEONSTORE_DB` unset.  Alternately, users can share a database as well by setting `NEONSTORE_DB` to a shared location, and adopting workflows which do ensure multiple users do not simultaneously call `neon_store()` to update the database.

@@

To download this data, `neon_download()` must make many API calls.  It first uses `neon_sites()` to reference all sites containing this data product.  Then for each site, for each month for which that site collected data, the function will make an API call to the `/data` endpoint (using the `neonstore` function, `neon_data()`), which returns a table of files found for that product at the requested site and month range.  The  table from `neon_data()` indicates 4 values: the file name, the MD5 or CRC32 hash of the data file (CRC32 hashes are used only for AOP data products), the file size (in bytes -- encoded as a character because some file sizes are too long to be represented as an integer), and a download URL.  The download URL embeds a temporary access token which expires after an hour, so users must complete all the downloads before this time expires.  The access string for each file is unique, thus it is (at the time of writing) impossible to download any given data file without making a corresponding API request, even if the precise file name and address is already known.  For the most part, this issue is invisible to `neonstore` users, as `neon_download()` calls automatically pass all download URLs to the download step.

`neon_download()` handles the following tasks:

- `neonstore` aims to provide persistent storage, writing raw data files to
  the appropriate app directory for your operating system (see `rappdirs`,
  [Ratnakumar et al 2016](https://CRAN.R-project.org/package=rappdirs)), now provided
  by the base R function `tools::R_user_dir()`.  Files are organized into appropriate 
  subdirectories by product, site, and year-month of sampling.  
- `neon_download()` provides  clean and concise progress bars for the two key
  processes involved: querying the API to obtain download URLs (which involves no 
  large data transfer but counts against API rate limiting, see below), and the
  actual file downloads.
- `neon_download()` will verify the integrity of file downloads against the MD5 or CRC32
  hashes provided. 
- `neon_download()` will omit downloads of any existing data files in the local 
  store (even if the timestamp portion of the filename has changed without updating the file)
- You can request multiple products at once using vector notation, though API
  rate limiting may interfere with large requests.
- `neon_download()` uses `curl::curl_download()` instead of `downloadr` package
  used in `neonUtilities`, which can be finicky on Windows and older versions of R. 
- Automatic handling NEON's rate limiting policy.  To do, `neon_download()` must  pause 100 seconds every 150 requests if an API token is not supplied or every 950 requests if a token is available.  Should a rate-limit error be detected, (possible if the same token is used in parallel requests), `neon_download()` will also pause for 100 seconds before retrying the failed request.   Requests and pauses are communicated through the user messaging interface.  
- Expanding compressed HDF5 files (`*h5.gz`). (Note: unusually, NEON chooses to compress binary h5 files with gzip, even though gzip compression is built internally into the H5 specification at configurable compression levels.  NEON does not choose to compress text-based formats such as .csv, where compression has larger gains.  Compression has implications for version management, where `neonstore` must track the hashes of compressed files to avoid duplicate downloads)
- Updating the local manifest in LMDB with the individual checksum and current release tag for each downloaded object (see Provenance)

```{r}
neon_download("DP1.10003.001", 
               start_date = "2018-01-01", 
               end_date = "2019-01-01",
               site = "YELL")
```

@@

Even adding a handful of products can generate thousands or hundreds of thousands of files in a local store. Surprisingly, these can even exceed the capacity of common `bash` shell utilities like `ls`.  `neonstore` sidesteps this issue through a convenient R-based interface.  The function `neon_index()` will list all recognized files found in the local store as a `data.frame`.  `neon_index()` uses a sophisticated filename parser, `neon_parse_filenames()`, that understands all documented file naming conventions listed at <https://data.neonscience.org/file-naming-conventions>, and quite a few that are not yet listed (internal NEON communication).  The most common of these are displayed in the table:

@@

`neon_index()` takes optional arguments such as `product`, `start_date`, `end_date`, and `site` to allow for convenient sub-setting.  Because most observational and instrument-based products are divided by site and month, this is still not a convenient way to browse the data.
